### 2025-06-28 paper/quarto_aaai/paper.tex
Add introduction to counterfactual generation, still at 7 pages

CEs have also been used to improve models in the natural language
processing (NLP) domain. First, \citet{wu2021polyjuice2} propose
\emph{Polyjuice}, a general-purpose counterfactual generator for
language models. The authors empirically demonstrate that the
augmentation of training data with their method improves robustness in a
number of NLP tasks. Second, \citet{balashankar2023improving} use
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves by up to 20\%. Finally,
\citet{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT) that aims to improve generalization and robustness of
language models by generating counterfactuals for training samples that
are subject to high predictive uncertainty.

@@

This section introduces the Counterfactual Training objective. CT
combines ideas from adversarial training, counterfactual explanations,
and energy-based modelling with the explicit goal of producing models
whose learned representations align with plausible explanations that
further comply with user-defined actionability constraints.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \citet{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see supplementary appendix.}

To train explainable models as defined in Def. \ref{def-explainability},
we propose to leverage counterfactuals in the following objective:

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (in 2\textsuperscript{nd} term) the contrastive divergence,
\(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and (in 3\textsuperscript{rd} term) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-off between the components can
governed through penalties \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

\noindent It is worth noting that \mbox{Proposition \ref{prp-mtblty}}
assumes independence of features. This raises a valid concern about the
effect of protecting immutable features in the presence of proxies that
remain unprotected. We address this in Section~\ref{sec-discussion}.

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
features: (1) the subjects' \emph{age}, which we define as immutable,
and (2) the subjects' existing level of \emph{debt}, which we define as
mutable. We have simulated this scenario using synthetic data with
independent \emph{age} and \emph{debt} features, and Gaussian
class-conditional densities in Figure~\ref{fig-poc}. The four panels
show the outcomes for different training procedures using the same model
architecture (a linear classifier). In panels (a) and (c) we have
trained the models conventionally, while in panels (b) and (d) we have
applied CT.

@@

\emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. In our experiments, grids with
270 settings took up to four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect: (1) CT yields itself to parallel execution, which
we have leveraged for our own experiments; (2) it amortizes the cost of
CEs for the training samples; (3) as demonstrated, it retains its value
when used as a ``fine-tuning'' technique for conventionally-trained
models.

\emph{Interventions on features have implications for fairness.} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features. Model owners can use our
solution to support the fair and equitable treatment of decision
subjects, but they could also misuse it by enforcing explanations based
on features that are more difficult to modify by some (group of)
individuals. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \citep{sharma2020certifai}, threatening the
equality of opportunity in the system \citep{bell2024fairness}.
Additionally, CT requires that model owners define the mutability
constraints for the features considered by the model. Even if all
immutable features are protected, there may exist proxies that are
mutable --- and hence should not be protected --- but preserve
sufficient information about the principals to hinder the protections.
Deciding on actionability is still a major open challenge in the AR
literature \citep{venkatasubramanian2020philosophical} impacting the
capacity of CT to fulfill its intended goal. As such, neither of these
phenomena is ultimately specific to CT.

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily discretized. Focus on
classification is a common choice in research on CEs and AR. Other
settings have attracted some interest, e.g., regression
\citep{spooner2021counterfactual}, but there is little consensus how to
robustly extend the notion of CEs.\\
Secondly, CT is susceptible to instabilities in training. This problem
has been recognized for JEMs \citep{grathwohl2020your} and even though
we depart from the SGLD-based sampling, we still encounter variability
in outcomes. CT is exposed to two potential sources of training
instabilities: (1) the energy-based contrastive divergence term in
Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
We find several promising ways to mitigate this problem: regularizing
energy (\(\lambda_{\text{reg}}\)), generating sufficiently many
counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures, and thus performance. In this work,
we have relied exclusively on grid searches for this task. Future work
on CT could benefit from investigating more sophisticated approaches.
Notably, our method is iterative which makes methods such as Bayesian or
gradient-based optimization applicable \citep{bischl2023hyperparameter}.

>>>>

CEs have also been used to improve models in the natural language
processing domain. For example, \citet{wu2021polyjuice2} propose
\emph{Polyjuice}, a general-purpose CE generator for language models and
demonstrate that the augmentation of training data with \emph{Polyjuice}
improves robustness in a number of tasks, while
\citet{luu2023counterfactual} introduce the \emph{Counterfactual
Adversarial Training} (CAT) framework that aims to improve
generalization and robustness of language models by generating
counterfactuals for training samples that are subject to high predictive
uncertainty.

@@

This section introduces the Counterfactual Training framework. CT
combines ideas from adversarial training, counterfactual explanations,
and energy-based modelling with the explicit goal of producing models
whose learned representations align with plausible explanations that
further comply with user-defined actionability constraints.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. In broad terms,
generating CEs using gradient descent entails solving some form of an
objective that balances (1) the classification loss
\(\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})\), and (2) one
or more penalty terms \(\lambda_{i}\text{cost}_{i}(\cdot)\). The exact
specification of these penalties induces various properties in the
counterfactual outcomes, and tends to be the key feature that
distinguishes various gradient-based ``generators'' or ``explainers''
(CE algorithms) from the literature \citep{altmeyer2023explaining}.

For our purposes, we let \(T\) vary and consider the search to be
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

CT leverages gradient-based CE search already in training to produce
perturbed versions of the factual samples \(\mathbf{x}\). Initially,
until \(\tau\) is reached or when \(t\leq T\), generated samples
\(\mathbf{x}_t^\prime\) serve as adversarial examples; we term them
\emph{nascent} counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Over
multiple training iterations up to time \(T\), these nascent
counterfactuals move towards the decision boundary and turn into
\emph{mature} counterfactuals \(\mathbf{x}_T^\prime\) when they cross
into the target class or when \(t > T\).

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
components are governed through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

\noindent Proposition \ref{prp-mtblty} assumes independent features,
which raises a concern about the effect of protecting immutable features
in the presence of proxies. We address it in
Section~\ref{sec-discussion}.

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
\(y\in\{1=\text{default},2=\text{no default}\}\) with two input
features: (1) the subjects' \emph{age}, which we define as immutable,
and (2) the subjects' existing level of \emph{debt}, which we define as
mutable. We have simulated this scenario using synthetic data with
independent \emph{age} and \emph{debt} features, and Gaussian
class-conditional densities in Figure~\ref{fig-poc}. The four panels
show the outcomes for different training procedures using the same model
architecture (a linear classifier). In panels (a) and (c) we have
trained the models conventionally, while in panels (b) and (d) we have
applied CT.

@@

\emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. Higher numbers of CEs
per iteration positively impact the quality of solutions, but they also
increase the amount of computations. In our experiments, grids with 270
settings took up to four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs (see supplementary
appendix for details). Three factors attenuate this effect: (1) CT
inherently yields itself to parallel execution; (2) it amortizes the
cost of CEs for the training samples; (3) it retains its value when used
to fine-tune conventionally-trained models.

\emph{Interventions on features have implications for fairness.} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features. Model owners can use our
solution to support the fair and equitable treatment of decision
subjects, but they could also misuse it by enforcing explanations based
on features that are more difficult to modify by some (group of)
individuals. When used irresponsibly, CT could result in an unfairly
assigned burden of recourse \citep{sharma2020certifai}, threatening the
equality of opportunity \citep{bell2024fairness}. Additionally, CT
requires mutability constraints for the features considered by the
model. Even if all immutable features are protected, there may exist
proxies that are mutable, and hence should not be protected, but
preserve sufficient information about the principals to hinder the
protections. Deciding on actionability is still a major open challenge
in the AR literature \citep{venkatasubramanian2020philosophical}
impacting the capacity of CT to fulfill its intended goal.

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily discretized. Focus on
classification is a common choice in research on CEs and AR. Other
settings have attracted some interest, e.g., regression
\citep{spooner2021counterfactual}, but there is little consensus how to
robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities. This problem has
been recognized for JEMs \citep{grathwohl2020your} and even though we
depart from the SGLD sampling, we still encounter variability in
outcomes. CT is exposed to two potential sources of instabilities: (1)
the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
Equation~\ref{eq-div}, and (2) the underlying explainers. We find
several promising ways to mitigate this problem: regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each epoch, and including only mature counterfactuals in
\(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures, and thus performance. In this work,
we have relied exclusively on grid searches for this task. Future work
on CT could benefit from investigating more sophisticated approaches.
Notably, our method is iterative which makes methods such as Bayesian or
gradient-based optimization applicable \citep{bischl2023hyperparameter}.

### 2025-06-28 paper/sections/discussion.qmd
Add introduction to counterfactual generation, still at 7 pages

*CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT yields itself to parallel execution, which we have leveraged for our own experiments; (2) it amortizes the cost of CEs for the training samples; (3) as demonstrated, it retains its value when used as a "fine-tuning" technique for conventionally-trained models.

*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features. Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Additionally, CT requires that model owners define the mutability constraints for the features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable --- and hence should not be protected --- but preserve sufficient information about the principals to hinder the protections. Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal. As such, neither of these phenomena is ultimately specific to CT.

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR. Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to instabilities in training. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter variability in outcomes. CT is exposed to two potential sources of training instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
Finally, we believe that it is possible to substantially improve hyperparameter selection procedures, and thus performance. In this work, we have relied exclusively on grid searches for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, our method is iterative which makes methods such as Bayesian or gradient-based optimization applicable [@bischl2023hyperparameter].

>>>>

*CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. Higher numbers of CEs per iteration positively impact the quality of solutions, but they also increase the amount of computations. In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see supplementary appendix for details). Three factors attenuate this effect: (1) CT inherently yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; (3) it retains its value when used to fine-tune conventionally-trained models.

*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features. Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. When used irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness]. Additionally, CT requires mutability constraints for the features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder the protections. Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR. Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD sampling, we still encounter variability in outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.  
Finally, we believe that it is possible to substantially improve hyperparameter selection procedures, and thus performance. In this work, we have relied exclusively on grid searches for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, our method is iterative which makes methods such as Bayesian or gradient-based optimization applicable [@bischl2023hyperparameter].

### 2025-06-28 paper/sections/method.qmd
Adjust experiments
Add introduction to counterfactual generation, still at 7 pages

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see supplementary appendix.]

<!-- @sec-app-ce -->

To train explainable models as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing (in 2\textsuperscript{nd} term) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (in 3\textsuperscript{rd} term) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can governed through penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

`\noindent`{=latex} It is worth noting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

## Example (Prediction of Consumer Credit Default)

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable. We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architecture (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.

>>>>

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. In broad terms, generating CEs using gradient descent entails solving some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$. The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining].

For our purposes, we let $T$ vary and consider the search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

CT leverages gradient-based CE search already in training to produce perturbed versions of the factual samples $\mathbf{x}$. Initially, until $\tau$ is reached or when $t\leq T$, generated samples $\mathbf{x}_t^\prime$ serve as adversarial examples; we term them *nascent* counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Over multiple training iterations up to time $T$, these nascent counterfactuals move towards the decision boundary and turn into *mature* counterfactuals $\mathbf{x}_T^\prime$ when they cross into the target class or when $t > T$.

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

`\noindent`{=latex} Proposition \ref{prp-mtblty} assumes independent features, which raises a concern about the effect of protecting immutable features in the presence of proxies. We address it in @sec-discussion.

## Example: Prediction of Credit Defaults

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable $y\in\{1=\text{default},2=\text{no default}\}$ with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable. We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architecture (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.
